---
title: "University of Washington Graduation Rate"
author: "James Lamar Foster"
date: "August 20, 2021"
output: html_document
---

```{r set_knitr_chunk_options, echo=FALSE, message=FALSE}
#set global output chunk options
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, collapse = F, highlight = F, results = "asis")
```

```{r}
# Packages for cleaning, wrangling, and visualizing
library(tidyverse)
library(janitor)
library(scales)
```

```{r}
# Read in the data 
ipeds_two <- read_csv("ipeds_final.csv")
```

```{r, eval=FALSE}
# Examine the data 
head(ipeds_two)
```

```{r}
#Clean variable names 
ipeds_two <- clean_names(ipeds_two)

#Rename Variables names 

ipeds_rename <- ipeds_two %>% 
  rename(
    total_enrollment = 3,
    total_bachelors_4 = 4, 
    total_bachelors_6 = 5,
    race_unknown_men = 6, 
    race_unknown_women = 7, 
    Male = 8, 
    Female = 9, 
    hispanic_men = 10, 
    hispanic_women = 11, 
    aian_men = 12, 
    aian_women = 13, 
    asian_men = 14, 
    asian_women = 15, 
    black_men = 16, 
    black_women = 17, 
    nhpi_men = 18, 
    nhpi_women = 19, 
    white_men = 20, 
    white_women = 21, 
    multi_men = 22, 
    multi_women = 23) %>% 
  select(-24)
```

```{r}
#Prepare data for visualization Graduate rate 
ipeds_grad_overall <- ipeds_rename %>% 
  select(2:5)

ipeds_grad_overall <- ipeds_grad_overall %>% 
  mutate(graduation_four_rate = round(total_bachelors_4/total_enrollment,2),
         graduation_six_rate = round(total_bachelors_6/total_enrollment,2),
         perc_four_rate = paste0(sprintf("%4.1f", total_bachelors_4 / total_enrollment * 100), "%"),
         perc_six_rate = paste0(sprintf("%4.1f", total_bachelors_6 / total_enrollment * 100), "%")) 
```

# **Summary of Findings**

*University of Washington's Graduate Rate*

- 4-year graduation rate is 44%
- 6-year graduation rate is 57%
- Female graduation rate is 38.4%
- Male graduation rate is 34.7%

This project uses data from the [Integrated Postsecondary Education Data System (IPEDS)](https://nces.ed.gov/ipeds/) to examine how the University of Washington - Seattle compares to similar institutions in the West regarding graduation. 

Four and six-year graduation rates were examined for the 2015 cohort. This exploratory analysis discovered that the University of Washington's 4-year and 6-year graduation rate ranks 5th among the 13 institutions in the Bureau of Economic Analysis (BEA) Far West region.

When the graduation rate was disaggregated by sex, the University of Washington's graduation rate decreased to sixth place for Females. However, the University of Washington remained in 5th place for Males, although the graduation percentage for males is lower than females' graduation percentage.

Analysts must do further disaggregation and analysis to understand the nuances of the University of Washington's graduation rate. For example, IPEDS provides Pell and race/ethnicity data, which may lend itself well to a deeper analysis. This may deepen our understanding of which groups of people need more support to help them graduate within 4 or 6 years.


**Note:**

*1. Graduation rate data within 150 percent of normal time - 4-year and 2-year institutions for female and male results*

*2. Number of adjusted cohort receiving a Bachelor's degree at 4 years (August 31, 2015)	First-time, full-time entering*

```{r}
#Visualize 4- year Graduate rate 
ipeds_grad_overall %>% 
  mutate(highlight = ifelse(institution_name == "University of Washington-Seattle Campus", T, F)) %>% 
  ggplot(aes(x = reorder(institution_name, graduation_four_rate), y = graduation_four_rate)) + 
  geom_col(aes(fill = highlight)) + 
  scale_fill_manual(values = c("grey60", "#4b2e83")) +
  coord_flip() +
  labs(title = "4-year Graduation Rate",
       subtitle = "For Far West Public Research Institutions",
       x = "",
       y = "",
       fill= "Institution Name",
       caption = "Data provided by Integrated Postsecondary Education Data System") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  geom_text(aes(label = perc_four_rate), hjust = 1, nudge_x = .1, color = "white", fontface = "bold" ) +
  scale_y_continuous(labels = scales::percent)
```



```{r}
# Visualize 6 year graduation rate 
ipeds_grad_overall %>% 
  mutate(highlight = ifelse(institution_name == "University of Washington-Seattle Campus", T, F)) %>% 
  ggplot(aes(x = reorder(institution_name, graduation_six_rate), y = graduation_six_rate)) + 
  geom_col(aes(fill = highlight)) + 
  scale_fill_manual(values = c("grey60", "#4b2e83")) +
  coord_flip() +
  labs(title = "6-year Graduation Rate",
       subtitle = "For Far West Public Research Institutions",
       x = "",
       y = "",
       fill= "Institution Name",
       caption = "Data provided by Integrated Postsecondary Education Data System") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  geom_text(aes(label = perc_six_rate), hjust = 1, nudge_x = .1, color = "white", fontface = "bold" ) +
  scale_y_continuous(labels = scales::percent)
```


```{r}
#Prepare data to visualize by Gender 
ipeds_gender <- ipeds_rename %>% 
  pivot_longer(cols = Male:Female, names_to = "sex", values_to = "total_enrolled") %>% 
  select(-(race_unknown_men:multi_women)) %>% 
   mutate(graduation_four_rate = round(total_bachelors_4/total_enrollment,4),
         graduation_six_rate = round(total_bachelors_6/total_enrollment,4),
         perc_four_rate = paste0(sprintf("%4.1f", total_bachelors_4 / total_enrollment * 100), "%"),
         perc_six_rate = paste0(sprintf("%4.1f", total_bachelors_6 / total_enrollment * 100), "%")) %>% 
  rename(Graduation = total_enrolled) %>% 
  select(-(graduation_four_rate:perc_six_rate)) %>% 
  select(-total_bachelors_4, -total_bachelors_6) %>% 
  mutate(Graduation_perc = round(Graduation/total_enrollment,4),
         perc_rate = paste0(sprintf("%4.1f", Graduation/total_enrollment * 100), "%")) 
```

```{r}
#Female Visualization 
ipeds_gender %>% 
  filter(sex == "Female") %>% 
  mutate(highlight = ifelse(institution_name == "University of Washington-Seattle Campus", T, F)) %>% 
  ggplot(aes(x = reorder(institution_name, Graduation_perc), y = Graduation_perc)) + 
  geom_col(aes(fill = highlight)) + 
  scale_fill_manual(values = c("grey60", "#4b2e83")) +
  coord_flip() +
  labs(title = "Female Graduation Rate",
       subtitle = "For Far West Public Research Institutions",
       x = "",
       y = "",
       fill= "Institution Name",
       caption = "Data provided by Integrated Postsecondary Education Data System") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  geom_text(aes(label = perc_rate), hjust = 1, nudge_x = .1, color = "white", fontface = "bold" ) +
  scale_y_continuous(labels = scales::percent)
```


```{r}
#Male Visualization 
ipeds_gender %>% 
  filter(sex == "Male") %>% 
  mutate(highlight = ifelse(institution_name == "University of Washington-Seattle Campus", T, F)) %>% 
  ggplot(aes(x = reorder(institution_name, Graduation_perc), y = Graduation_perc)) + 
  geom_col(aes(fill = highlight)) + 
  scale_fill_manual(values = c("grey60", "#4b2e83")) +
  coord_flip() +
  labs(title = "Male Graduation Rate",
       subtitle = "For Far West Public Research Institutions",
       x = "",
       y = "",
       fill= "Institution Name",
       caption = "Data provided by Integrated Postsecondary Education Data System") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  geom_text(aes(label = perc_rate), hjust = 1, nudge_x = .1, color = "white", fontface = "bold" ) +
  scale_y_continuous(labels = scales::percent)
```

```{r}
ipeds_gender %>% 
  mutate(highlight = ifelse(institution_name == "University of Washington-Seattle Campus", T, F)) %>% 
  ggplot(aes(x = reorder(institution_name, Graduation_perc), y = Graduation_perc)) + 
  geom_col(aes(fill = highlight)) + 
  scale_fill_manual(values = c("grey60", "#4b2e83")) +
  coord_flip() +
  labs(title = "Graduation Rate",
       subtitle = "For Far West Public Research Institutions",
       x = "",
       y = "",
       fill= "Institution Name",
       caption = "Data provided by Integrated Postsecondary Education Data System") +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  geom_text(aes(label = perc_rate), hjust = 1, nudge_x = .1, color = "white", fontface = "bold" ) +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~sex)
```








